The Influence of Connectivity on Dung Beetle Communities

### 2025-07-07 corridor_docs/eric_ms_thesis/MS_thesis.Rmd
holy canoli

# TO DO LIST

FOR TOMORROW:

1)  Finish adding citations

2)  Stat results: biology first, then stats (e.g., abundance ranged form xxx in rectangular to xxx in winged)

3)  Review other papers to find:

-   those with similar results
-   those with contrary results
-   do this with dung bettles AND other insect taxa

DOWN THE ROAD

1)  **Results:** Did you get any species not captured in other long-leaf/SE USA forest sites?

2)  **Methods:** Find me the best example paragraph you think you read describing the use of hill numbers to estimate diversity/richness (field survey paper)

3)  **Methods:** Find me the best TWO descriptions of using GLMM to compare abundance/diversity/richness at the Corridor Project

4)  **Methods:** Write a short paragraph on how you calculated and compared functional diversity (how you measured it, how you compared it)

5)  **Figure:** biomass of different (R/T/D) functional groups.

6)  **Table:** Results of functional diversity analyses.

@@

As human disturbances continue to expand into natural landscapes, intact habitats are becoming increasingly fragmented [@diaz_pervasive_2019; @ma_global_2023; @taubert_global_2018]. Like many ecological processes, fragmentation is a complex and multifaceted phenomenon bringing about many consequences which can be both positive and negative for ecosystems [@fahrig_effects_2003; @fletcher_is_2018]. However, as habitats are broken down community structures are significantly altered [@WOS:000419965700012; @haddad_corridor_2003; @jennings_composition_2006; **harrison and bruna**]. This alteration of structure typically lends to loss in biodiversity on a global scale and interruptions in ecosystem processes and functions [@haddad_habitat_2015].

@@

*[**EB edit:** break this next sentence into 3 to carry reader though the logic: changes in species diverstiy are important to understand because species are involved in interactions. If we gain or lose species we could gain or lose interactions and resulting ecosystems processes. there is some evidence of this, but... look at the green text notes below]*

@@

By measuring changes in biodiversity and species richness within experimental designs we are able to isolate factors might be contributing to ecological patterns and processes [@fletcher_jr_landscape_2023; @resasco_contribution_2017]. Past studies have endeavored to experimentally measure changes in community compositions as a result of connecting habitats with corridors [@collins_fragmentation_2017; @graham_impact_2022; @tewksbury_corridors_2002] **here fix citations**, yet very few have directly compared matrix and patch populations. **examples of dispersal concepts (drift fence etc** Land use is different from one species to another so it is vital to understand where compositions are distributed and what processes might be driving population differences.

@@

Dung beetles have emerged as a model system with which to test spatial ecology hypotheses [@ros_how_2012; @roslin_dung_2000].They are an incredibly well studied group of insects which are well known for driving a multitude of ecosystem functions [@hasan_dung_2024]. The removal, breakdown, and burial of animal feces drive important ecosystem interactions provided by dung beetles enhancing nutrient cycling and soil quality, the reduction of breeding sites for parasites, and a reduction in methane emissions from dung [@slade_disentangling_2016; @WOS:000350444300013; @nichols_ecological_2008]. [**EB edit:** I added the following to justify dung beetles more, and then made the questions a second paragraph] Local assemblages of dung beetles can be species-rich with species comprising a broad range of functional traits [e.g., size, foraging style, resource-use, [@decastro-arrazola_trait-based_2023; @ospina-garces_dung_2018]]. Previous studies have shown that isolated patches of habitat frequently have lower dung beetle diversity and abundance than areas of continuous habitat, as well as documented their presence in linear strips of habitat that resemble corridors [@WOS:000707546700001]. Past studies have also focused on how landscape structure alters the community compositions of dung beetles[@WOS:000398662200001], yet large landscape scale experimental studies with carefully controlled and replicated treatments are non-existent for this model species.

@@

\begin{flushleft}  
\textbf{OUTLINE}
\end{flushleft}

@@

In our comparison of dung beetle compositions, connectivity and habitat edge accounted for differences in dung beetle abundances, yet species richness and diversity were even across patch types and in  the matrix. Total beetle counts were consistently highest in the matrix, in comparison abundances is connected patches were not significantly different while those of the winged and rectangular patches were lower. Additionally, we also observed that patch effects were not equally proportional for all species. This suggests that the species captured could be exhibiting habitat preferences between the open patches versus forested matrix. Another potential explanation is that species in the matrix are acting as a source population which feeds into patches with edge acting as a drift fence for directing movement. The latter example seems more likely because flies, which have similar resource dependencies were also found to have a similar interaction ** fried **, and also studies on dung beetles in the tropics found more dramatic differences in dung beetle populations in fragments and matrix [@bustamante-sanchez_dung_2004; @WOS:000180797500010; @WOS:000288810500026].

@@

Although richness and diversity were the same among treatments, there was notable variability between sampling blocks. Block 8 generally had the lowest species richness and biodiversity while 53n had the highest. While the design of the experimental blocks control for as much as possible, landscape features across the sight are different enough to influence species composition. Several factors may explain this pattern, including soil quality which influences where beetles can reproduce [@arellano_response_2008; @conover_phenology_2019] as well as matrix permeability, canopy cover, and different land-use types (e.g., roadways, water sources)** harvey **[@WOS:000651446900001]. Not only do these impact dung beetle movement, but also these landscape features can affect mammal movement, which in turn limits dung availability for beetles.

@@

Despite dung beetles being a well studied model system, most of the current work in fragmented habitats come from tropical climates[@nichols_global_2007], with any work in temperate locations coming from Europe where _Aphodius_ beetles dominate[@roslin_distribution_2001]. Many studies in the tropics have found patterns of habitat preference between open fields and continuous forest similar to what we did in this study[@arellano_response_2008; @WOS:000349242000004]. However, differences in composition were much more distinctly different with some species being exclusive to habitat types unlike we see in this paper. Of the limited work out of the south eastern United States we did recover assemblages similar to other work [@conover_phenology_2019; @nealis_habitat_1977; @young_pines_nodate], and even returned results to another study where differences between treatment types were non-significant while variation on a larger scale was more apparent[@young_dung_2023]. In comparison to other studies from SRS our results were similar to thise  of small mammals which were more likely to be using matrix land *mabry*. We did expect some species to be more dominant in patches with higher edge ratios as was found with ants at SRS [@resasco_landscape_2014], but species were evenly present in all patches.

@@

- potential effects of seasonality

- Sampling wasn't designed to look for potential effects of edge proximity, and so we couldn't test for previously observed pattern of generalist/opportunistic species more common near edges.

- used great bait, but possible some species (eg specialists) not attracted

@@

- tough to compare when most of the work has been done in the tropics; surprisingly few studies of this species-rich and ecolgically vital guild in the SE US. "Despite ample work documenting their ecological importance, there is a suprising lack of research on dung beetle diversity and corridors outside of the tropics.  That said, tropical studies find stark contrast between fragments/pasture. They have been found to use corridors, but it is...."

- experimental approaches to track movement and distinguish beetles already in pacthes vs. those coming from outside. Our preliminary data with MRR indicates individuals do move through corridors.

@@

-   dung beetles are robust

-   implications for ecosystem processes.

-   what do think about corridor design and considerations for fragmented landscapes

-   practical applications think about the beetles

\begin{flushleft} 
\textbf{A few things to consider incorporating}
\end{flushleft}

- ellen wind paper/ connectivity

This study advances our understanding of the factors shaping dung beetle community composition in temperate regions of the southeastern United States. In addition, the experimental design enables direct comparisons between populations in continuous matrix habitat and those in both isolated and corridor connected patches. Our main findings emphasized: (1) Habitat type and patch shape were the main driving factors for dung beetle species abundances were composed, however effects were species specific. (2) Patch shape and isolation had less of an influence on species richness which was relatively even on both a patch and block level. (3) Species diversity metrics were also relatively even across patch types however varied widely by sampling blocks. These results suggest that dung beetle species are fully capable of permeating fragmented landscapes and that habitat type and connectivity shape community compositions, but landscape effects of a larger scale are driving changes in biodiversity.

@@

>>>>

@@

As human disturbances continue to expand into natural landscapes, intact habitats are becoming increasingly fragmented [@diaz_pervasive_2019; @ma_global_2023; @taubert_global_2018]. Like many ecological processes, fragmentation is a complex and multifaceted phenomenon bringing about many consequences which can be both positive and negative for ecosystems [@fahrig_effects_2003; @fletcher_is_2018]. However, as habitats are broken down community structures are significantly altered [@WOS:000419965700012; @haddad_corridor_2003; @jennings_composition_2006; [@harrison_habitat_1999]]. This alteration of structure typically lends to loss in biodiversity on a global scale and interruptions in ecosystem processes and functions [@haddad_habitat_2015].

@@

@@

By measuring changes in biodiversity and species richness within experimental designs we are able to isolate factors might be contributing to ecological patterns and processes [@fletcher_jr_landscape_2023; @resasco_contribution_2017]. Past studies have endeavored to experimentally measure changes in community compositions as a result of connecting habitats with corridors [@collins_fragmentation_2017; @graham_impact_2022; @tewksbury_corridors_2002]  Yet very few have directly compared matrix and patch populations. Land use is different from one species to another so it is vital to understand where compositions are distributed and what processes might be driving population differences[@haddad_corridor_1999].

@@

Dung beetles have emerged as a model system with which to test spatial ecology hypotheses [@ros_how_2012; @roslin_dung_2000].They are an incredibly well studied group of insects which are well known for driving a multitude of ecosystem functions [@hasan_dung_2024]. The removal, breakdown, and burial of animal feces drive important ecosystem interactions provided by dung beetles enhancing nutrient cycling and soil quality, the reduction of breeding sites for parasites, and a reduction in methane emissions from dung [@slade_disentangling_2016; @WOS:000350444300013; @nichols_ecological_2008]. Local assemblages of dung beetles can be species-rich with species comprising a broad range of functional traits [e.g., size, foraging style, resource-use, [@decastro-arrazola_trait-based_2023; @ospina-garces_dung_2018]]. Previous studies have shown that isolated patches of habitat frequently have lower dung beetle diversity and abundance than areas of continuous habitat, as well as documented their presence in linear strips of habitat that resemble corridors [@WOS:000707546700001]. Past studies have also focused on how landscape structure alters the community compositions of dung beetles[@WOS:000398662200001], yet large landscape scale experimental studies with carefully controlled and replicated treatments are non-existent for this model species.

@@

@@

In our comparison of dung beetle compositions, connectivity and habitat edge accounted for differences in dung beetle abundances, yet species richness and diversity were even across patch types and in  the matrix. Total beetle counts were consistently highest in the matrix, in comparison abundances is connected patches were not significantly different while those of the winged and rectangular patches were lower. Additionally, we also observed that patch effects were not equally proportional for all species. This suggests that the species captured could be exhibiting habitat preferences between the open patches versus forested matrix. Another potential explanation is that species in the matrix are acting as a source population which feeds into patches with edge acting as a drift fence for directing movement. The latter example seems more likely because flies, which have similar resource dependencies were also found to have a similar interaction [@fried_habitat_2005], and also studies on dung beetles in the tropics found more dramatic differences in dung beetle populations in fragments and matrix [@bustamante-sanchez_dung_2004; @WOS:000180797500010; @WOS:000288810500026].

@@

Although richness and diversity were the same among treatments, there was notable variability between sampling blocks. Block 8 generally had the lowest species richness and biodiversity while 53n had the highest. While the design of the experimental blocks control for as much as possible, landscape features across the sight are different enough to influence species composition. Several factors may explain this pattern, including soil quality which influences where beetles can reproduce [@arellano_response_2008; @conover_phenology_2019] as well as matrix permeability, canopy cover, and different land-use types (e.g., roadways, water sources)[@WOS:000241362400032; @WOS:000651446900001]. Not only do these impact dung beetle movement, but also these landscape features can affect mammal movement, which in turn limits dung availability for beetles.

@@

Despite ample work documenting their ecological importance, there is a surprising lack of research on dung beetle diversity and corridors outside of the tropics[@nichols_global_2007], with any work in temperate locations coming from Europe where _Aphodius_ beetles dominate[@roslin_distribution_2001]. Many studies in the tropics have found patterns of habitat preference between open fields and continuous forest similar to what we did in this study[@WOS:000349242000004], and some work that found that forest species were using living fences as corridors[@arellano_response_2008]. However, differences in composition were much more distinctly different with some species being exclusive to habitat types unlike we see in this paper. Of the limited work out of the south eastern United States we did recover assemblages similar to other work [@conover_phenology_2019; @nealis_habitat_1977; @young_pines_nodate], and even returned results to another study where differences between treatment types were non-significant while variation on a larger scale was more apparent[@young_dung_2023]. In comparison to other studies from SRS our results were similar to those of small mammals which were more likely to be using matrix land *mabry*. We did expect some species to be more dominant in patches with higher edge ratios as was found with ants at SRS [@resasco_landscape_2014], but species were evenly present in all patches.

@@

There were some limitations of this study which should be addressed in any future work to capture a clearer picture of how populations are being altered by habitat connectivity. First, sampling for this study was conducted over a two month period in the summer of 2024, sufficient data was collected but due to a lack of available resources temporal patterns were obscured due to inconsistent sampling periods. This is highly important since dung beetles exhibit consistent patterns of seasonality [@conover_phenology_2019; @davis_feeding_1966]. Another potential avenue for improvement is lowering the grain size of sampling and changing trap placement to better understand the effects of edge proximity and connectivity. We also used one of the best bait types for collecting dung beetles in pig dung [@marsh_optimising_2013], but a mix of differently sourced baits may have been more optimal as more diverse baits attract more diverse species [@frank_nutrient_2017; @frank_search_2018; @gimenez_gomez_dung_2021].

@@

Moving forward, there is a bounty of knowledge yet to be collected for dung beetles in the south eastern United States. Relative to what is already known in tropical ecosystems, this study provides a glimpse on how dung beetles are responding to fragmentation and connectivity in subtropic pine habitats, but much can still be learned about dung beetles in these spaces and the role they play in ecosystems as well as how they interact with landscapes. The unique role of dung beetles as waste removers is fascinating enough and of great importance for maintaining landscapes so dominated by pastures like Southern US [@cheng_dweller_2022; @WOS:000368243300001]. Not only waste removal but also secondary effect such as seed dispersal and parasitic reduction are important to understand in the context of fragmentation [@fincher_effects_1975; @vulinec_dung_2002]. SRS provides an excellent experimental design for direct comparison of landscapes but another main goal of the site is to study dispersal, dung beetles would be an excellent system for studying movement. They already show promise, in preliminary trials we released beetles from both the connected and rectangular patches with baited traps in the central patch. We recovered one recapture which originated from the connected patch, indicating that individuals do move through corridors.

@@

Although there is plentiful work to be done, this study paints an encouraging picture that while dung beetles face disturbances that make habitats less preferential, they are robust enough to persist throughout fragmented landscapes and the management strategies we do have to connect fragmented landscapes work to mitigate losses. Because assemblages remain mostly the same across these landscapes, ecosystem services should remain uninterrupted allowing continuing benefits to all community members. While this study of dung beetles in fragmented landscapes was conducted in a relatively protected area, in real world application we could consider the use of movement corridors within disturbed areas to help bolster effected populations, but also near areas of high importance such as pastures to provide areas of refuge for beetles. In this case it is important to consider the inverse of what was manipulated in this study where open field acts as matrix and forested area is focal landscape, thus it is vital to continue learning more about how landscape composition and connectivity effect dung beetle populations.

@@

This study advances our understanding of the factors shaping dung beetle community composition in subtropical regions of the southeastern United States. In addition, the experimental design enables direct comparisons between populations in continuous matrix habitat and those in both isolated and corridor connected patches. Our main findings emphasized: (1) Habitat type and patch shape were the main driving factors for dung beetle species abundances were composed, however effects were species specific. (2) Patch shape and isolation had less of an influence on species richness which was relatively even on both a patch and block level. (3) Species diversity metrics were also relatively even across patch types however varied widely by sampling blocks. These results suggest that dung beetle species are fully capable of permeating fragmented landscapes and that habitat type and connectivity shape community compositions, but landscape effects of a larger scale are driving changes in biodiversity.
